%  =========================================================================
%  David P. Fleming's PhD Dissertation - University of Washington, Department of Astronomy, 2020
%  =========================================================================

%  =========================================================================
%   University of Washington PhD Thesis Template
%   Copyright (c) 1985 The University of Washington
%
%  Licensed under the Apache License, Version 2.0 (the "License");
%  you may not use this file except in compliance with the License.
%  You may obtain a copy of the License at
%
%      http://www.apache.org/licenses/LICENSE-2.0
%
%  Unless required by applicable law or agreed to in writing, software
%  distributed under the License is distributed on an "AS IS" BASIS,
%  WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
%  See the License for the specific language governing permissions and
%  limitations under the License.
%  ========================================================================
%
 
\documentclass [11pt, proquest]{uwthesis} % draft

\usepackage{amsmath,amssymb}
\usepackage{aasmacros}
\usepackage[version=4]{mhchem}
\usepackage{comment}
\usepackage{multirow,tabularx}
\usepackage{deluxetable}
\usepackage{mathtools,upgreek}
\usepackage{floatrow}
\usepackage{outlines}
\usepackage{setspace}
\usepackage{caption}
\usepackage{sidecap}
\captionsetup[table]{font={stretch=1.0}}     %% change 1.2 as you like
\captionsetup[figure]{font={stretch=1.0}}    %% change 1.2 as you like
\usepackage{epigraph}
\usepackage{times}
\usepackage{mathptmx}
\usepackage{bibentry}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage[round]{natbib}
\usepackage{xcolor}
\usepackage{import}
\usepackage{pythonhighlight}

\PassOptionsToPackage{hyphens}{url}
 \usepackage[
     pdfauthor={David P. Fleming},
    pdftitle={Inferring the Evolutionary Histories of Stars and Their Planets},
    pdfsubject={Astronomy},
    pdfcreator={pdflatex},
    bookmarks=true,         % show bookmarks bar?
     pdfnewwindow=true,      % links in new window
     colorlinks=true,    % false: boxed links; true: colored links
     linkcolor=xlinkcolor,     % color of internal links
     citecolor=xlinkcolor,     % color of links to bibliography
     filecolor=xlinkcolor,  % color of file links
     urlcolor=xlinkcolor,      % color of external links
     pdfpagemode=UseOutlines,
    final=true
    ]{hyperref}
\usepackage{hypcap}

%% Custom commands
\def\mearth{{\rm\,M_\oplus}}
\def\rearth{{\rm\,R_\oplus}}
\def\msun{{\rm\,M_\odot}}
\def\rsun{{\rm\,R_\odot}}
\def\lsun{{\rm\,L_\odot}}
\def\gsim{~\rlap{$>$}{\lower 1.0ex\hbox{$\sim$}}}
\def\lsim{~\rlap{$<$}{\lower 1.0ex\hbox{$\sim$}}}
\newcommand{\xxx}[1]{{\textbf{#1}}}
\newcommand{\vplanet}[0]{\texttt{VPLanet}\xspace}
\newcommand{\emcee}[0]{\texttt{emcee}\xspace}
\newcommand{\approxposterior}[0]{\texttt{approxposterior}\xspace}
\newcommand{\eqtide}[0]{\texttt{EQTIDE}\xspace}
\newcommand{\stellar}[0]{\texttt{STELLAR}\xspace}
\newcommand{\atmesc}[0]{\texttt{AtmEsc}\xspace}
\newcommand{\kepler}[0]{\textit{Kepler}\xspace}
\newcommand{\tess}[0]{\texttt{TESS}\xspace}
\newcommand{\jwst}[0]{\textit{JWST}\xspace}
\newcommand{\binary}[0]{\texttt{BINARY}\xspace}
\newcommand{\rebound}[0]{\texttt{REBOUND}\xspace}
\newcommand{\um}{$\upmu$m}

%% for editing changes
% hyperref link defaults to "blue" (0000ff) as this matches our publisher produced pdf style
\definecolor{xlinkcolor}{cmyk}{0.5,0.5,0,1}

\def\bibpreamble{\protect\addcontentsline{toc}{chapter}{Bibliography}}

\setcounter{tocdepth}{2}  % Print the chapter and sections to the toc

\begin{document}
 
% ==========   Preliminary pages
%
% ( revised 2012 for electronic submission )
%

\prelimpages
 
%
% ----- Copyright and title pages
%
\Title{Inferring the Evolutionary Histories of Stars and Their Planets}
\Author{David P. Fleming}
\Year{2020}
\Program{Astronomy}

\Chair{Rory Barnes}{Professor}{Astronomy}
\Signature{Tom Quinn}
\Signature{Victoria Meadows}
\Signature{Jacob VanderPlas}

% May be first or second. Must follow specified order. 
\titlepage  

\newpage

% May be first or second.
\copyrightpage

%
% ----- Abstract
%

\include{tex/abstract}

%
% ----- Contents & etc.
%

\tableofcontents
\listoffigures
\listoftables  
 
%
% ----- Acknowledgments
%
\acknowledgments{% \vskip2pc
  % {\narrower\noindent
I am incredibly lucky to have been supported by many fantastic people throughout my life. The gratitude I offer pales in comparison to what I owe these people for everything they have done for me. Thank you to my wonderful wife, Anya, and adorable kitten, Bandit, for their endless love and support. Thank you to my parents, Tim and Anne, and my siblings, Jack and Kristin. Without their care and support, none of this would be possible. Thank you to my great friends, Jake and Emily, that helped make Seattle home. Thank you to all the Astronomy graduate students, past and present, that have helped me through courses, qualifying exams, W(h)inetimes, and in countless other ways. Thank you to the Astronomy Department faculty, my advisor Rory, Tom, Jake VanderPlas, and Vikki for their teachings, patience, support, and for setting an excellent example. Thank you to the entire Virtual Planetary Laboratory team for your friendship, support, and awesome collaborations. Finally, I would like to acknowledge that we are on the traditional land of the first people of Seattle, the Duwamish People past and present and honor with gratitude the land itself and the Duwamish Tribe.
  % \par}
}
%
% ----- Dedication
%
\dedication{\begin{center}To my wife, Anya\end{center}}

%
% end of the preliminary pages
 
%
% ==========      Text pages
%

\textpages
 
% ========== Chapters

% Intro
\chapter{Introduction}
\import{tex/}{intro.tex}

% CB disks
\chapter{The Resonant Gravitational Interactions Between Binary Stars and Circumbinary Protoplanetary Disks}
\import{tex/CBDisk/}{cbdisk.tex}

% STEEP
\chapter{The Dynamical Death of Circumbinary Planets: Coupled Stellar-Tidal Evolution Ejection of Circumbinary Exoplanets}
\import{tex/STEEP/}{bin_tides.tex}

% Sync
\chapter{The Competition between Magnetic Braking and Tidal Torques in Binary Stars: Application to the \kepler Field}
\import{tex/sync/}{sync.tex}

% approxposterior
\chapter{Efficient Approximate Bayesian Inference with \approxposterior}
\import{tex/approxposterior/}{approxposterior.tex}

% TRAPPIST-1
\chapter{Inferring the XUV History of the TRAPPIST-1 Planetary System}
\import{tex/trappist/}{trappist.tex}

% Conclusions
\chapter{Conclusions}
\import{tex/}{conclusions.tex}

\printendnotes

%
% ==========   Bibliography
%
%\nocite{*}   % include everything in the uwthesis.bib file
\bibliographystyle{apalike} % plain if not natbib
\bibliography{thesisBib}
%
% ==========   Appendices
%

\appendix
\raggedbottom\sloppy

\end{document}
